---
title: "Covid-19 forecast model exploration"
subtitle: "Exploring and visualizing COVID-19 forecast models"
date: "Last updated: `r format(Sys.time(),'%B %d, %Y')`"
output: 
  html_document:
    toc: false
    toc_float: false
    df_print: paged
    code_download: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width=12, fig.height=6)
```

```{r, echo=FALSE, include=FALSE}
# clear environment
rm(list = ls())  # clear memory
```

Load required packages
```{r, message=FALSE}
# load pacman package to load or install other requried libraries
if (!require('pacman')) install.packages('pacman'); library(pacman)

# load (install if required) packages from CRAN
p_load("here","tidyverse","lubridate","janitor","plotly","doParallel")

# load/install packages from GitHub
p_load_gh("reichlab/zoltr", "reichlab/covidHubUtils")
```

Utilizing `zoltr` package to enable access to Zoltar API to the forecast archive and `covidHubUtils` package to that provide functions to read, plot and score forecast data.

\  

### Importing observed data
We will use the Covid-19 cases as updated by CDC. We will download data from the data Table for daily Case Trends for The United States [found here]( https://covid.cdc.gov/covid-data-tracker/#trends_dailycases). 

Also, we will pull daily COVID-19 cases from CDC at a state-level [found here](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36/data).

Data extracted September, 06, 2021. Here is the top rows of the incident COVID-19 cases data.
```{r, message=FALSE}
# csv export of daily cases from CDC COVID data tracker at national level
national_cases_daily <- read_csv(here("data", "data_table_for_daily_case_trends__the_united_states.csv"), 
                                 col_types = cols(Date = col_date(format = "%b %d %Y")),
                                 skip = 2)


# csv export of daily cases from CDC COVID data tracker at state level
states_cases_daily <- read_csv(here("data", "United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv"), 
                                        col_types = cols(submission_date = col_date(format = "%m/%d/%Y")))

head(national_cases_daily)
```

\  

### Preparing data

We'll aggregate the daily case counts into weeks, starting on Sunday as is the same as an epidemiological week (referred to as MMWR week, more info [found here](https://ndc.services.cdc.gov/wp-content/uploads/MMWR_week_overview.pdf))
```{r}
# clean up column names
national_cases_daily <- national_cases_daily %>%
  clean_names()

# aggregate daily COVID cases into weekly case counts, 
# start week on Sunday according to epidemiological week (MMWR week) 
week <- as.Date(cut(national_cases_daily$date, "week", start.on.monday = FALSE))
national_cases_weekly <- aggregate(new_cases ~ week, national_cases_daily, sum)
head(national_cases_weekly)

# filter dates after August
national_cases_weekly <- national_cases_weekly %>%
  filter(week < "2021-09-01")
```

The forecast models are pulled from the [Zoltar forecast model archive](https://www.zoltardata.com/) using the [covidHubUtils package](http://reichlab.io/covidHubUtils/index.html). 

```{r, message=FALSE}
# check which US-specific models are contained in the forecast archive.
model_names <- get_all_models(hub = "US") 
```

There are `r length(model_names)` models within the Zoltar forecast archive from the U.S COVID-19 Forecasts Hub.

Load the incident cases forecasts models of 1 through 4 week horizons of a select number of US-specific models that are published by the CDC and are contained within the Zoltar forecast archive.

```{r, include=FALSE}
# include all forecast models or a few specified models
load_all_models <- 0
```

```{r, message=FALSE, cache=TRUE}
if (load_all_models == 1) {
# load forecasts of all models  
system.time(all_model_inc_case_targets <- load_forecasts(
  location = "US",
  hub = "US",
  types = c("point","quantile"),
  targets =  paste(1:4, "wk ahead inc case"),
  as_of = "2021-09-06",
  source = "zoltar"))
  }
```

```{r, message=FALSE, cache=TRUE}
# load a select few point forecasts that were submitted in from zoltar forecast archive
  inc_case_targets <- load_forecasts(
  models = c("COVIDhub-ensemble", "CovidAnalytics-DELPHI", "CU-select",
             "IHME-CurveFit", "LANL-GrowthRate","USC-SI_kJalpha", 
             "JHU_IDD-CovidSP", "UVA-Ensemble"),
  location = "US",
  hub = "US",
  types = c("point","quantile"),
  targets =  paste(1:4, "wk ahead inc case"),
  as_of = "2021-09-06",
  source = "zoltar")
```

```{r, include=FALSE}
# the number of models with week incident case forecasts
if (load_all_models == 1){
  length(unique(all_model_inc_case_targets$model))
  } else {
    length(unique(inc_case_targets$model))
  }
```

54 models stored in the Zoltar forecast archive have submitted a weekly incident case forecast.

```{r}
# display top rows of forecast data
head(inc_case_targets)

inc_case_targets <- inc_case_targets %>%
  mutate(forecast_day = lubridate::wday(forecast_date, label = TRUE, abbr = FALSE)) %>%
  select(model, forecast_date, forecast_day, everything())
```

```{r, include=FALSE}
# adjust forecast dates to lie on the same week of their respective forecast target end date
# https://github.com/reichlab/covid19-forecast-hub/blob/master/data-processed/README.md#forecast-file-format
adj_inc_case_targets <- inc_case_targets %>%
  mutate(adj_forecast_date = as.Date(ifelse(forecast_day == "Sunday", 
                                    forecast_date,
                                    case_when(forecast_day == "Monday" ~ forecast_date - 1,
                                              forecast_day == "Tuesday" ~ forecast_date + 5,
                                              forecast_day == "Wednesday" ~ forecast_date + 4,
                                              forecast_day == "Thursday" ~ forecast_date + 3,
                                              forecast_day == "Friday" ~ forecast_date + 2,
                                              forecast_day == "Saturday" ~ forecast_date + 1)), 
                                    origin = "1970-01-01"),
         adj_forecast_day = lubridate::wday(adj_forecast_date, label = TRUE, abbr = FALSE)) %>%
  select(forecast_date, forecast_day, adj_forecast_date, adj_forecast_day, target_end_date, everything()) %>%
  arrange(adj_forecast_date) %>%
  # dplyr::distinct(model, adj_forecast_date, .keep_all = TRUE) %>%
  filter(adj_forecast_date < "2021-09-01")
```

Filter the models for respective 1 through 4 week horizon point forecasts.
```{r}
# filter for 1-4 week horizon point forecasts
for (wk in 1:4) {
  wk_forecasts <- adj_inc_case_targets %>% 
  filter(horizon == wk,
         type == "point") %>%
  select(-quantile) %>% 
  dplyr::distinct(model, adj_forecast_date, .keep_all = TRUE)
  assign(paste0("inc_case_targets_",wk,"_week"), wk_forecasts)
}

# inc_case_targets_1_week <- adj_inc_case_targets %>% 
#   filter(horizon == 1,
#          type == "point") %>%
#   select(-quantile)
```

### Visualizing forecast data

Plot the CDC actual COVID-19 cases versus the forecast predictions for the select models.
```{r}
# plot the weekly cases for United States according to CDC data as of 09/06/2021
p <- ggplot(data = national_cases_weekly, aes(x = week, y = new_cases)) +
  geom_point() +
  geom_line() +
  labs(title = "CDC weekly incident COVID-19 cases")
p
```

```{r, include=FALSE}
# plot various model 1 week horizon forecasts
ggplot(data = inc_case_targets_1_week, aes(x = adj_forecast_date, y = value, color = model)) +
  geom_point() +
  geom_line() +
  theme(legend.position = "bottom")
```

```{r}
# combine plots
p + geom_line(data = inc_case_targets_1_week, aes(x = target_end_date, y = value, color = model)) +
  geom_point(data = inc_case_targets_1_week, aes(x = target_end_date, y = value, color = model)) +
  labs(title = "CDC weekly incident COVID-19 cases versus various model point forecasts") +
  theme(legend.position = "bottom")
```

Interactive plot of CDC observed cases versus 1 week horizon forecasts
```{r}
fig1 <- plot_ly(type = 'scatter',  mode = 'lines+markers')
fig1 <- fig1 %>% 
  add_trace(data = inc_case_targets_1_week, x = ~adj_forecast_date, y = ~value, color = ~factor(model)) %>%
  add_trace(data = national_cases_weekly, x = ~week, y = ~new_cases, name = "CDC actual", color = I('black')) %>%
  layout(title = "CDC weekly incident COVID-19 cases\n and various model point forecasts")

fig1
```

IHME-Curvefit only predicted incident COVID-cases for a few weeks before focusing on predicting hospitalizations and deaths, according to predictions stored in Zoltar forecast archive.

Identifying peaks of COVID-19 cases
```{r}
# a 'peak' is defined as a local maxima with m points either side of it being smaller than it. hence, the bigger the parameter m, the more stringent is the peak finding procedure
# https://github.com/stas-g/findPeaks
find_peaks <- function (x, m = 3){
    shape <- diff(sign(diff(x, na.pad = FALSE)))
    pks <- sapply(which(shape < 0), FUN = function(i){
       z <- i - m + 1
       z <- ifelse(z > 0, z, 1)
       w <- i + m + 1
       w <- ifelse(w < length(x), w, length(x))
       if(all(x[c(z : i, (i + 2) : w)] <= x[i + 1])) return(i + 1) else return(numeric(0))
    })
     pks <- unlist(pks)
     pks
}
```


Peaks of observed weekly incident COVID-19 cases based on CDC data
```{r}
# find peaks of observed cases
peak_cases <- national_cases_weekly[find_peaks(national_cases_weekly$new_cases, m = 3),]

# plot peaks for observed covid cases
q <- p + geom_point(data = peak_cases, aes(x = week, y = new_cases, color = "CDC actual")) +
  labs(title = "Observed peaks of weekly incident COVID-19 cases", 
       x = "Weeks", y = "Incident cases", color = "Peaks") +
  theme(legend.position = "bottom")

q
```

```{r}
# get peak values of all values
model_peaks_list <- list()
for (i in unique(inc_case_targets_1_week$model)) {
  models <- inc_case_targets_1_week %>%
    filter(model == i)
  peaks <- models[find_peaks(models$value, m = 3),]
  model_peaks_list[[i]] <- peaks
}

# combine the df of peaks
model_peaks <- bind_rows(model_peaks_list)
```

Look at the peaks associated with forecast models
```{r}
# combine plots of cdc actual peaks with forecast model peaks
q + geom_point(data = model_peaks, aes(x = adj_forecast_date, y = value, color = model))
```

```{r}
# find magnitude and temporal differences in the peaks
cdc_peaks <- peak_cases %>%
  mutate(model = rep("CDC", nrow(peak_cases))) %>%
  select(week, model, new_cases) %>%
  rename("peaks" = "new_cases")

model_observed_pks <- model_peaks %>%
  select(adj_forecast_date, model, value) %>%
  rename("week" = "adj_forecast_date",
         "peaks" = "value") %>%
  bind_rows(cdc_peaks)

```

```{r, include=FALSE}
# find earliest peak for each of the models
model_observed_pks %>%
  group_by(model) %>%
  summarise(first_pk_date = min(week))
```

### Score each model weekly
To score models using covidHubUtils, data frame needs to be in same format as one gotten from `load_truth` function.
```{r, message=FALSE, cache=TRUE}
# load a truth data frame from covidHub
jhu_truth_df <- load_truth(
  truth_source = c("JHU"),
  target_variable = c("inc case"),
  truth_end_date = "2021-09-04",
  temporal_resolution = "weekly",
  hub = "US",
  locations =  "US"
)

# top rows of truth dataframe
head(jhu_truth_df)
```

```{r}
# to score models, data frame needs to be in same format as one gotten from load_truth
# TODO: convert national_cases_weekly to jhu_truth_df format
# TODO: use score_forecast function to compute weighted interval score
```


