---
title: "Covid-19 forecast model exploration"
subtitle: "Exploring and visualizing COVID-19 forecast models"
date: "Last updated: `r format(Sys.time(),'%B %d, %Y')`"
output: 
  html_document:
    toc: false
    toc_float: false
    df_print: paged
    code_download: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width=12, fig.height=6)
```

```{r, echo=FALSE, include=FALSE}
# disable scientific notation
options(scipen = 999, digits = 2)

# clear environment
rm(list = ls())  # clear memory
```

Load required packages
```{r, message=FALSE}
# load pacman package to load or install other requried libraries
if (!require('pacman')) install.packages('pacman'); library(pacman)

# load (install if required) packages from CRAN
p_load("here","tidyverse","lubridate","janitor","data.table","plotly","doParallel")

# load/install packages from GitHub
p_load_gh("reichlab/zoltr", "reichlab/covidHubUtils")
```

Utilizing `zoltr` package to enable access to Zoltar API to the forecast archive and `covidHubUtils` package to that provide functions to read, plot and score forecast data.

\  

### Importing observed data
We will use the Covid-19 cases as updated by CDC. We will download data from the data Table for daily Case Trends for The United States [found here]( https://covid.cdc.gov/covid-data-tracker/#trends_dailycases). 

Also, we will pull daily COVID-19 cases from CDC at a state-level [found here](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36/data).

Data extracted September, 06, 2021. Here is the top rows of the incident COVID-19 cases data.
```{r, message=FALSE}
# csv export of daily cases from CDC COVID data tracker at national level
national_cases_daily <- read_csv(here("data", "data_table_for_daily_case_trends__the_united_states.csv"), 
                                 col_types = cols(Date = col_date(format = "%b %d %Y")),
                                 skip = 2)


# csv export of daily cases from CDC COVID data tracker at state level
states_cases_daily <- read_csv(here("data", "United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv"), 
                                        col_types = cols(submission_date = col_date(format = "%m/%d/%Y")))

head(national_cases_daily)
```

\  

### Preparing data

We'll aggregate the daily case counts into weeks, starting on Sunday as is the same as an epidemiological week (referred to as MMWR week, more info [found here](https://ndc.services.cdc.gov/wp-content/uploads/MMWR_week_overview.pdf))
```{r}
# clean up column names
national_cases_daily <- national_cases_daily %>%
  clean_names()

# aggregate daily COVID cases into weekly case counts, 
# start week on Sunday according to epidemiological week (MMWR week) 
week <- as.Date(cut(national_cases_daily$date, "week", start.on.monday = FALSE))
national_cases_weekly <- aggregate(new_cases ~ week, national_cases_daily, sum)
head(national_cases_weekly)

# include last day of week and filter dates after August
national_cases_weekly <- national_cases_weekly %>%
  mutate(week_end = week + days(6)) %>%
  filter(week < "2021-09-01")
  
```


The forecast models are pulled from the [Zoltar forecast model archive](https://www.zoltardata.com/) using the [covidHubUtils package](http://reichlab.io/covidHubUtils/index.html). 

```{r, message=FALSE}
# check which US-specific models are contained in the forecast archive.
model_names <- get_all_models(hub = "US") 
```

There are `r length(model_names)` models within the Zoltar forecast archive from the U.S COVID-19 Forecasts Hub.

Load the incident cases forecasts models of 1 through 4 week horizons of a select number of US-specific models that are published by the CDC and are contained within the Zoltar forecast archive.

```{r, include=FALSE}
# bypass querying all forecast models form zoltar
bypass_load_from_zoltar <- 1
```

```{r, message=FALSE, cache=TRUE}
if (bypass_load_from_zoltar == 0) {
# load forecasts of all models  
system.time(all_inc_case_targets <- load_forecasts(
  location = "US",
  hub = "US",
  types = c("point","quantile"),
  targets =  paste(1:4, "wk ahead inc case"),
  as_of = "2021-09-06",
  source = "zoltar"))
  
# write model forecast to directory
write_csv(all_inc_case_targets, here("data", "all_inc_case_targets.csv"))
} else {
  # read pre-queried forecast data
  all_inc_case_targets <- read_csv(here("data", "all_inc_case_targets.csv"))
}

```

```{r, message=FALSE, cache=TRUE}
# filter for  a select few point forecasts that were submitted in from zoltar forecast archive
  inc_case_targets <- all_inc_case_targets %>% 
  filter(model %in% c("COVIDhub-ensemble", "CovidAnalytics-DELPHI", "CU-select",
             "IHME-CurveFit", "LANL-GrowthRate","USC-SI_kJalpha", 
             "JHU_IDD-CovidSP", "UVA-Ensemble"))

```

```{r, include=FALSE}
# the number of models with week incident case forecasts
  length(unique(all_inc_case_targets$model))

  length(unique(inc_case_targets$model))
```

52 models stored in the Zoltar forecast archive have submitted a weekly incident case forecast.

```{r}
# display top rows of forecast data
head(inc_case_targets)

inc_case_targets <- inc_case_targets %>%
  mutate(forecast_day = lubridate::wday(forecast_date, label = TRUE, abbr = FALSE)) %>%
  select(model, forecast_date, forecast_day, everything())
```

```{r, include=FALSE}
# adjust forecast dates to lie on the same week of their respective forecast target end date
# https://github.com/reichlab/covid19-forecast-hub/blob/master/data-processed/README.md#forecast-file-format
adj_inc_case_targets <- inc_case_targets %>%
  mutate(adj_forecast_date = as.Date(ifelse(forecast_day == "Sunday", 
                                    forecast_date,
                                    case_when(forecast_day == "Monday" ~ forecast_date - 1,
                                              forecast_day == "Tuesday" ~ forecast_date + 5,
                                              forecast_day == "Wednesday" ~ forecast_date + 4,
                                              forecast_day == "Thursday" ~ forecast_date + 3,
                                              forecast_day == "Friday" ~ forecast_date + 2,
                                              forecast_day == "Saturday" ~ forecast_date + 1)), 
                                    origin = "1970-01-01"),
         adj_forecast_day = lubridate::wday(adj_forecast_date, label = TRUE, abbr = FALSE)) %>%
  select(forecast_date, forecast_day, adj_forecast_date, adj_forecast_day, target_end_date, everything()) %>%
  arrange(adj_forecast_date) %>%
  # dplyr::distinct(model, adj_forecast_date, .keep_all = TRUE) %>%
  filter(adj_forecast_date < "2021-09-01")
```

Filter the models for respective 1 through 4 week horizon point forecasts.
```{r}
# filter for 1-4 week horizon point forecasts
for (wk in 1:4) {
  wk_forecasts <- adj_inc_case_targets %>% 
  filter(horizon == wk,
         type == "point") %>%
  select(-quantile) %>% 
  dplyr::distinct(model, adj_forecast_date, .keep_all = TRUE)
  assign(paste0("inc_case_targets_",wk,"_week"), wk_forecasts)
}

```

### Visualizing forecast data

Plot the CDC actual COVID-19 cases versus the forecast predictions for the select models.
```{r}
# plot the weekly cases for United States according to CDC data as of 09/06/2021
p <- ggplot(data = national_cases_weekly, aes(x = week_end, y = new_cases)) +
  geom_point() +
  geom_line() +
  labs(title = "CDC weekly incident COVID-19 cases")
p
```

```{r, include=FALSE}
# plot various model 4 week horizon forecasts
ggplot(data = inc_case_targets_4_week, aes(x = target_end_date, y = value, color = model)) +
  geom_point() +
  geom_line() +
  theme(legend.position = "bottom")
```

```{r}
# combine plots
p + geom_line(data = inc_case_targets_4_week, aes(x = target_end_date, y = value, color = model)) +
  geom_point(data = inc_case_targets_4_week, aes(x = target_end_date, y = value, color = model)) +
  labs(title = "CDC weekly incident COVID-19 cases versus various model point forecasts") +
  theme(legend.position = "bottom")
```

Interactive plot of CDC observed cases versus 4 week horizon forecasts
```{r, include=FALSE}
plot_inc_forecast <-  function(horizon = 1) {
  fig1 <- plot_ly(type = 'scatter',  mode = 'lines+markers')
  fig1 <- fig1 %>% 
    add_trace(data = get((paste0("inc_case_targets_",horizon,"_week"))), x = ~target_end_date, y = ~value, color = ~factor(model)) %>%
    add_trace(data = national_cases_weekly, x = ~week_end, y = ~new_cases, name = "CDC actual", color = I('black')) %>%
    layout(title = paste("CDC weekly incident COVID-19 cases\n and", horizon, "week horizon point forecasts"))
  fig1
}

```

```{r}
# plot 1 week horizon point forecast
plot_inc_forecast(horizon = 1)

# plot 4 week horizon point foreacst
plot_inc_forecast(horizon = 4)
```

IHME-Curvefit only predicted incident COVID-cases for a few weeks before focusing on predicting hospitalizations and deaths, according to predictions stored in Zoltar forecast archive.

Identifying peaks of COVID-19 cases
```{r}
# a 'peak' is defined as a local maxima with m points either side of it being smaller than it. hence, the bigger the parameter m, the more stringent is the peak finding procedure
# https://github.com/stas-g/findPeaks
find_peaks <- function (x, m = 3){
    shape <- diff(sign(diff(x, na.pad = FALSE)))
    pks <- sapply(which(shape < 0), FUN = function(i){
       z <- i - m + 1
       z <- ifelse(z > 0, z, 1)
       w <- i + m + 1
       w <- ifelse(w < length(x), w, length(x))
       if(all(x[c(z : i, (i + 2) : w)] <= x[i + 1])) return(i + 1) else return(numeric(0))
    })
     pks <- unlist(pks)
     pks
}
```


Peaks of observed weekly incident COVID-19 cases based on CDC data
```{r}
# find peaks of observed cases
peak_cases <- national_cases_weekly[find_peaks(national_cases_weekly$new_cases, m = 3),]

# plot peaks for observed covid cases
q <- p + geom_point(data = peak_cases, aes(x = week_end, y = new_cases, color = "CDC actual")) +
  labs(title = "Observed peaks of weekly incident COVID-19 cases", 
       x = "Weeks", y = "Incident cases", color = "Peaks") +
  theme(legend.position = "bottom")

q
```

```{r}
# get peak values of all values
model_peaks_list <- list()

# get peaks for 4 week horizon forecast
for (i in unique(inc_case_targets_4_week$model)) {
  models <- inc_case_targets_4_week %>%
    filter(model == i)
  peaks <- models[find_peaks(models$value, m = 3),]
  model_peaks_list[[i]] <- peaks
}

# combine the df of peaks
model_peaks <- bind_rows(model_peaks_list)
```

Look at the peaks associated with forecast models
```{r}
# combine plots of cdc actual peaks with forecast model peaks
q + geom_point(data = model_peaks, aes(x = target_end_date, y = value, color = model))
```

```{r}
# find magnitude and temporal differences in the peaks
cdc_peaks <- peak_cases %>%
  mutate(model = rep("CDC", nrow(peak_cases))) %>%
  select(week_end, model, new_cases) %>%
  rename("peaks" = "new_cases")

model_observed_pks <- model_peaks %>%
  select(target_end_date, model, value) %>%
  rename("week_end" = "target_end_date",
         "peaks" = "value") %>%
  bind_rows(cdc_peaks)

```

```{r, include=FALSE}
# # find earliest peak for each of the models
# model_observed_pks %>%
#   group_by(model) %>%
#   summarise(first_pk_date = min(week_end))
```

### Score each model weekly
To score models using covidHubUtils, data frame needs to be in same format as one gotten from `load_truth` function.
```{r, message=FALSE, cache=TRUE}
# load truth data frame from file
bypass_truth_zoltar <- 1

# load a truth data frame from covidHub
if (bypass_truth_zoltar == 0) {
  jhu_truth_df <- load_truth(
    truth_source = c("JHU"),
    target_variable = c("inc case"),
    truth_end_date = "2021-09-04",
    temporal_resolution = "weekly",
    hub = "US",
    locations =  "US")
  
  # write truth data frame to csv file
  write_csv(jhu_truth_df, here("data", "jhu_truth_df.csv"))
  } else {
    
    # load truth df from file
    jhu_truth_df <- read_csv(here("data", "jhu_truth_df.csv"))
    }

# top rows of truth dataframe
head(jhu_truth_df)
```

```{r, include=FALSE}
# to score models, data frame needs to be in same format as one gotten from load_truth
# TODO: convert national_cases_weekly to jhu_truth_df format

# filter dates after September 6, 2021
jhu_truth_df2 <- jhu_truth_df %>% 
  filter(target_end_date < "2021-09-06")

# join CDC observed cases and rearrange truth dataframe
jhu_truth_df2 <- jhu_truth_df2 %>%
  full_join(national_cases_weekly, by = c("target_end_date" = "week_end")) %>%
  select(jhu_model = model, jhu_value = value, value = new_cases, everything())

# add name for CDC observed data and rearrange
truth_df <- jhu_truth_df2 %>%
  mutate(model = rep("Observed data (CDC)", nrow(jhu_truth_df2))) %>%
  select(model, everything()) %>%
  select(-jhu_model, -jhu_value)

#use score_forecasts function to compute weighted interval score and other metrics
forecast_model_scores <- score_forecasts(forecasts = all_inc_case_targets, 
                                          truth = truth_df, 
                                          metrics = c("abs_error", "wis", "wis_components", 
                                                      "interval_coverage"))

```


```{r, include=FALSE}
# join forecast df to get forecast values for each model
joint_df <- all_inc_case_targets %>%
  filter(type == "point") %>%
  right_join(forecast_model_scores, by = c("model", "horizon","forecast_date","target_end_date")) %>% 
  dplyr::rename("forecast_value" = value)
```

WIS scores for each model and different forecast horizons
```{r}
# get each models weekly wis score for each horizon
for (i in 1:4) {
  wis_scores <- joint_df %>% 
    filter(horizon == i) %>%
    # group_by(model, target_end_date) %>%
    select(model, horizon, abs_error, wis, forecast_value, true_value, target_end_date, forecast_date) %>%
    arrange(desc(wis))
  
  assign(paste0("wis_scores_",i), wis_scores)
}
wis_scores_1
wis_scores_2
wis_scores_3
wis_scores_4
```

Median absolute percent error (mape) for each model and forecast horizon
```{r, message=FALSE}
# calculate the median absolute percent error for each model
joint_df %>%
  group_by(model, horizon) %>%
  summarise(num_of_forecasts = n(),
            mape = median(abs((true_value-forecast_value)/true_value))*100) %>%
  arrange(desc(mape))
```

```{r}
# filter out dates before July 2020, labor day, thanksgiving, Christmas and New Years
national_cases_weekly %>%
  filter(week_end > "2020-07-04" & 
           !week_end == "2020-09-12" & 
           !week_end == "2020-12-26" & 
           !week_end == "2021-01-02")
  


# check the weeks of monotonic increasing and decreasing cases time periods
weekly_cases <- data.table(national_cases_weekly)
weekly_cases <- weekly_cases %>%
  mutate(diff = new_cases - lag(new_cases, 1, 0),
         inc = if_else(diff>0, 1, 0),
         dec = if_else(diff<0, 1, 0),
         inc_dec_count = sequence(rle(as.character(inc))$lengths))

# consecutive weeks of increasing COVID cases 
weekly_cases[which(inc==1), ]

# consecutive weeks of increasing COVID cases 
weekly_cases[which(inc==0), ]


# check the weeks of monotonic increasing and decreasing cases time periods for each model
models_inc_dec <- all_inc_case_targets %>%
  filter(type == "point",
         horizon == 4) %>%
  select(-location, -temporal_resolution, -target_variable, -type, -quantile, -location_name, 
         -population, -starts_with("geo"), -abbreviation, -full_location_name) %>%
  group_by(model) %>%
  mutate(diff = value - lag(value, 1,0),
         inc = if_else(diff>0, 1, 0),
         dec = if_else(diff<0, 1, 0),
         inc_dec_count = sequence(rle(as.character(inc))$lengths))

models_inc_dec
```

```{r}
# check the weeks of monotonic increasing and decreasing cases time periods for each model
models_horizon_inc_dec <- all_inc_case_targets %>%
  filter(type == "point") %>%
  select(-location, -temporal_resolution, -target_variable, -type, -quantile, -location_name, 
         -population, -starts_with("geo"), -abbreviation, -full_location_name) %>%
  group_by(model, horizon) %>%
  mutate(diff = value - lag(value, 1,0),
         inc = if_else(diff>0, 1, 0),
         dec = if_else(diff<0, 1, 0),
         inc_dec_count = sequence(rle(as.character(inc))$lengths))

models_horizon_inc_dec
```





